Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding Across Toronto’s Neighborhoods

### 2024-11-27 paper/paper.qmd
Update Paper

---
title: "How Urban Pollinator Garden Types and Locations Influence Community-Led Grant Success in Toronto"
subtitle: "Insights from Bayesian Analysis on Funding Outcomes for PollinateTO Projects."
author: 
  - Aliza Abbas Mithwani
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
```

@@

## Telegraphing
The remainder of this paper is structured as follows. Section 2 describes the data used in the analysis, including its sources, key features, and considerations regarding measurement and visualization. Section 3 outlines the methodology, detailing the data visualization techniques employed and the justification for using a Bayesian logistic regression model to analyze the relationship between garden characteristics and funding outcomes. Section 4 presents the results, including model validation and key findings. Section 5 offers a discussion of the results, focusing on their implications through key points, followed by an examination of the study's limitations and suggested future steps for further exploration. Section 6 concludes the paper. The Appendix provides supplementary materials, including raw data previews, data feature definitions, data visualizations, and technical model details such as posterior predictive checks, MCMC convergence diagnostics, and additional figures.

@sec-data....

# Data {#sec-data}
## Software Packages
We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
## Data Source

## Data Features

## Data Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Data Consideration

## Methodology

## Data Visualization

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

#analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

#analysis_data |> 
#  ggplot(aes(x = width, y = length)) +
#  geom_point(alpha = 0.8) +
#  theme_minimal() +
#  labs(x = "Wing width (mm)",
#       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project.

>>>>

---
title: "Key Drivers of Community Garden Growth: Analyzing Trends in PollinateTO Funding Across Toronto’s Neighborhoods"
subtitle: "A Bayesian Approach to Understanding the Impact of Time, Type, and Location on Garden Size and Number."
author: 
  - Aliza Abbas Mithwani
thanks: "Code and data are available at: [https://github.com/alizamithwani/PollinateTO.git]"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

#### Workspace setup ####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(arrow)
```

@@

## Telegraphing
The remainder of this paper is structured as follows. @sec-data describes the data used in the analysis, including its sources, key features, and considerations regarding measurement and visualization. Section 3 outlines the methodology, detailing the data visualization techniques employed and the justification for using a Bayesian logistic regression model to analyze the relationship between garden characteristics and funding outcomes. Section 4 presents the results, including model validation and key findings. Section 5 offers a discussion of the results, focusing on their implications through key points, followed by an examination of the study's limitations and suggested future steps for further exploration. Section 6 concludes the paper. The Appendix provides supplementary materials, including raw data previews, data feature definitions, data visualizations, and technical model details such as posterior predictive checks, MCMC convergence diagnostics, and additional figures.

# Data {#sec-data}
## Software Packages
We use the statistical programming language R [@citeR], rstanarm [@rstanarm], tidyverse [@tidyverse] Our data [@PTdata].... Following @tellingstories, we consider...

## Data Source
This study uses data from the PollinateTO Primary Project Garden Locations dataset, hosted on Toronto's Open Data Portal. The dataset provides information on community-led garden projects funded by PollinateTO, including geographic location, garden type, funding year, and other features relevant to understanding the distribution and characteristics of these initiatives. It is publicly accessible and serves as a vital resource for analyzing PollinateTO’s impact in Toronto neighborhoods.

## Data Features
The dataset includes several key variables:

YEAR_FUNDED: The year the project received funding from PollinateTO.
GARDEN_TYPE: Categories of gardens (e.g., food gardens, pollinator gardens, rain gardens).
NIA_OR_EN: Classification of garden location as part of a Neighbourhood Improvement Area (NIA), Emerging Neighborhood (EN), or other areas.
WARD_NAME: The city ward where the garden is located.
IS_INDIGENOUS_GARDEN: A binary variable indicating whether the garden is Indigenous-led.
PROJECT_NAME: Names assigned to specific garden projects.
LATITUDE and LONGITUDE: Geographic coordinates of each garden.
These features allow for diverse analyses, from exploring spatial trends to assessing equity and accessibility in funding distribution.

## Data Measurement*** (check key)
The dataset translates real-world community garden projects into structured entries by standardizing key attributes. For example:

YEAR_FUNDED records when projects were initiated, reflecting the temporal dimension of PollinateTO’s funding priorities.
GARDEN_TYPE is based on project applications, categorized to highlight garden purposes and ecological roles.
NIA_OR_EN and WARD_NAME capture the geographic and socioeconomic contexts of gardens, derived from city planning frameworks.
IS_INDIGENOUS_GARDEN identifies projects promoting Indigenous cultural or ecological knowledge, based on project self-identification during funding application.
The dataset provides structured snapshots of projects, enabling analysis of broader trends, equity impacts, and ecological contributions of urban greening initiatives.

## Data Consideration
Several considerations affect the dataset’s completeness and reliability:

Self-Reported Data: Much of the data, including garden types and Indigenous-led status, relies on self-reporting by project applicants.
Selection Bias: Funding decisions prioritize certain areas (e.g., NIAs), potentially overrepresenting them compared to other regions.
Temporal Incompleteness: The dataset spans 2019 to present, limiting analyses to recent funding trends.
Granularity: Certain variables, such as project outcomes or specific plant species, are not included, restricting the scope of ecological impact analysis.
Geographic Precision: While coordinates are provided, some garden boundaries and precise sizes may not be captured, introducing potential spatial analysis challenges.

## Methodology
Several considerations affect the dataset’s completeness and reliability:

Self-Reported Data: Much of the data, including garden types and Indigenous-led status, relies on self-reporting by project applicants.
Selection Bias: Funding decisions prioritize certain areas (e.g., NIAs), potentially overrepresenting them compared to other regions.
Temporal Incompleteness: The dataset spans 2019 to present, limiting analyses to recent funding trends.
Granularity: Certain variables, such as project outcomes or specific plant species, are not included, restricting the scope of ecological impact analysis.
Geographic Precision: While coordinates are provided, some garden boundaries and precise sizes may not be captured, introducing potential spatial analysis challenges.

## Data Visualization
To understand the data provided better, we visualize it using tables and graphs. To begin with, let's see how many gardens each type of neighborhood (NIA, EN or None) has and separate them by whether or not they are indigenous-led in @fig-IndNeiBoxPlot and @fig-IndNeiBarChart.

```{r}
#| label: fig-IndNeiBoxPlot
#| fig-cap: Summary of Gardens by Neighborhood Type
#| echo: false

#### Read data ####
PT_analysis_data <- read_parquet("../data/02-analysis_data/PT_analysis_data.parquet")

# Summarize the data
PT_summary_table <- PT_analysis_data %>%
  group_by(nia_or_en) %>%
  summarise(
    Total_Gardens = n(), # Total number of gardens in each neighborhood type
    Indigenous_Gardens = sum(is_indigenous_garden), # Count of Indigenous-led gardens
    Non_Indigenous_Gardens = Total_Gardens - Indigenous_Gardens, # Count of Non-Indigenous gardens
    Proportion_Indigenous = (Indigenous_Gardens / Total_Gardens) * 100, # Proportion of Indigenous gardens
    .groups = "drop"
  )

# Display the table
PT_summary_table %>%
  kable(
    col.names = c(
      "Neighborhood Type (NIA or EN)", 
      "Total Gardens", 
      "Indigenous-Led Gardens", 
      "Non-Indigenous Gardens", 
      "Proportion Indigenous (%)"
    )
  )
```

```{r}
#| label: fig-IndNeiBarChart
#| fig-cap: Distribution of Garden Types Funded Over Time
#| echo: false
# Filter for Indigenous-Led gardens only
PT_indigenous <- PT_analysis_data %>%
  filter(is_indigenous_garden == 1) %>%
  group_by(nia_or_en) %>%
  summarise(count = n(), .groups = "drop")

# Create a bar chart
ggplot(PT_indigenous, aes(x = nia_or_en, y = count, fill = nia_or_en)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Count of Indigenous-Led Gardens by Neighborhood Type",
    x = "Neighborhood Type (NIA or EN)",
    y = "Number of Indigenous-Led Gardens",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

Next, we visualize the trend of garden sizes and numbers across time to examine the impact of the initiative over time in @fig-trendline.

```{r}
#| label: fig-trendline
#| fig-cap: Trend of Garden Size and Number Over Time
#| echo: false

# Summarize the data by year to get the total number of gardens and average garden size per year
PT_summary <- PT_analysis_data %>%
  group_by(year_funded) %>%
  summarise(
    total_gardens = sum(number_of_gardens, na.rm = TRUE), # Summing number of gardens per year
    avg_garden_size = mean(estimated_garden_size, na.rm = TRUE) # Calculating average garden size
  )

# Create the line chart with two y-axes (total gardens and average size)
ggplot(PT_summary, aes(x = year_funded)) +
  geom_line(aes(y = total_gardens), color = "blue", linewidth = 1.2, linetype = "solid") + 
  geom_line(aes(y = avg_garden_size), color = "red", linewidth = 1.2, linetype = "dashed") +
  scale_y_continuous(
    name = "Total Number of Gardens", 
    sec.axis = sec_axis(~., name = "Average Garden Size (Estimated)")
  ) +
  labs(
    x = "Year Funded",
    y = "Total Number of Gardens"
  ) +
  theme_minimal() +
  theme(
    axis.title.y.right = element_text(color = "red"),
    axis.text.y.right = element_text(color = "red")
  )
```
\newpage
Now, let's analyze the proportion of each garden type visually across time in @fig-distributiontype to see if certain garden types increased in number and proportion overtime, reflecting shifting funding priorities and initiatives.

```{r}
#| label: fig-distributiontype
#| fig-cap: Distribution of Garden Types Funded Over Time
#| echo: false

# Summarize the data to get the count of garden types by year
PT_summary <- PT_analysis_data %>%
  group_by(year_funded, garden_type) %>%
  summarise(
    garden_count = n(), # Count the number of occurrences for each garden type per year
    .groups = "drop"
  )

# Create the bar chart
ggplot(PT_summary, aes(x = year_funded, y = garden_count, fill = garden_type)) +
  geom_bar(stat = "identity", position = "stack") + # Stacked bar chart
  scale_fill_brewer(palette = "Set3") + # Use a color palette for clarity
  labs(
    x = "Year Funded",
    y = "Number of Gardens",
    fill = "Garden Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 1) # Rotate x-axis labels for readability
  )
```
\newpage
To prevent outliars from making graphs unreadable, we use log-transformed garden sizes to analyze differences in garden sizes based on neighborhood type in @fig-neighborhoodsize. A boxplot is used to do this where the 3 neighborhood categories are EN (Emerging Neighborhoods), NIA (Neighborhood Improvement Areas) and None (neither EN nor NIA).

```{r}
#| label: fig-neighborhoodsize
#| fig-cap: Log-Transformed Garden Sizes by Neighborhood Type
#| echo: false
# Add a log-transformed column for garden size
PT_analysis_data <- PT_analysis_data %>%
  mutate(log_garden_size = log1p(estimated_garden_size)) # log1p handles zero values safely

# Use the log-transformed column in the boxplot
ggplot(PT_analysis_data, aes(x = nia_or_en, y = log_garden_size, fill = nia_or_en)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, notch = FALSE) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Neighborhood Type (NIA or EN)",
    y = "Log of Estimated Garden Size",
    fill = "Neighborhood Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```
# Model
In the analysis, I used a Bayesian Logistic Regression Model to examine the relationship between garden size and other factors like the type of garden, the year the project was funded, the neighborhood type (emerging neighborhood, EN vs neighborhood improvement area, NIA), and if it is an indigenous-led project.
